Dynamics of Laterally Propagating Flames in X-ray Bursts. I. Burning Front Structure

### 2020-02-18 xrb/submit/paper.tex

%\documentclass[12pt, preprint]{aastex}
\documentclass[preprint,times,tighten]{aastex63}

@@

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods, which approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
us to understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere as well
as the nature of any turbulence the burning front might encounter as it propagates
through the atmosphere.

@@

The computational expense of the problem necessitated several approximations
to make this set of simulations \citep{cavecchi:2013, art-2015-cavecchi-etal, art-2016-cavecchi-etal, Cavecchi2019}
tractable. The first was the assumption of vertical hydrostatic equilibrium, which
allowed them to filter out vertically propagating sound waves that would otherwise
have limited the timestep. This enabled them to reach the simulation times (on the
order of seconds) needed to capture the rise of the burst, but they could not model
vortical fluid motions driven by effects such as turbulence and convection. These can
facilitate mixing and help to accelerate the propagation of the flame front. They also
included only the triple-$\alpha$ process in their implementation of nuclear burning,
and use a fixed, constant opacity across the entire domain for each simulation run (although
they explore the dependence of the flame propagation speed on this opacity value).
Our simulations relax the assumption of vertical hydrostatic equilibrium, and implement
a more complete set of nuclear reactions and opacities. However, as discussed later, this
comes at the expense of a different set of approximations.

@@

The temperature profile in the star and ramp region is given as:
\begin{equation}
T(y) = T_\star + \frac{1}{2} (T_\mathrm{hi} - T_\star) \left [ 1 + \tanh\left( \frac{\tilde{y}}{2 \delta_\mathrm{atm}} \right ) \right ]
\end{equation}
with
\begin{equation}
\tilde{y} = y - H_\star - \frac{3}{2} \delta_\mathrm{atm}
\end{equation}
Here, $\delta_\mathrm{atm}$ is a characteristic width of the
transition ramp, $T_\star$ is the temperature of the neutron star, and
$T_\mathrm{hi}$ is the highest temperature in the HSE model---it will
represent the base of the fuel layer. We also define a floor temperature,
$T_\mathrm{lo}$, which the atmosphere reaches and is held constant at in
the outer, isothermal layer.

@@

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, ${X_k}_\mathrm{atm}$, and $g$.  We
then integrate outwards from the base of the ramp region ($y =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we find $p_i$ and
$\rho_i$ using a Newton-Raphson solver together with the equation of state, with either the temperature
specified, $T_i = T(p_i, \rho_i, \{X_k\}_i)$ (in the isothermal, ramp,
and buffer layers) or constant entropy, $s_i = s(p_i, \rho_i,
\{X_k\}_i)$, in the fuel layer.  This follows the procedures described in \citet{ppm-hse}.  We
use a constant temperature for all $y < H_\star +
3\delta_\mathrm{atm}$.  Above this, we switch to an isentropic atmosphere until the
temperature drops to $T_\mathrm{lo}$, at which point we
again keep the temperature constant.  The integration of the
atmosphere continues until the density falls to a low density cutoff,
$\rho_\mathrm{cutoff}$.  The material above this height is taken to
have constant density and temperature.
The choice of factors in front of $\delta_\mathrm{atm}$ were designed
to make sure the peak $T$ is attained at the desired density
of the burning layer.
The parameters we use for
the model generation are listed in Table~\ref{table:params} and the 
initial model profiles are showing in Figure~\ref{fig:initial_models}.

@@

To measure the propagation rate of the burning front, we first collapse our nuclear
energy generation rate ($\dot{e}_\mathrm{nuc}$) data at each time into a 1D radial
profile by averaging over the vertical coordinate. We then take the peak
$\dot{e}_\mathrm{nuc}$ value across all profiles to provide a fixed reference point.
We define the position of the flame front to be the location to the right of hottest part of
the flame where $\dot{e}_\mathrm{nuc}$ first drops to $< 0.1 \%$ of the global maximum. 
This corresponds roughly to the leading edge of the burning region. Averaging
over the vertical coordinate helps to reduce sensitivity to localized fluid motions, as does
tracking the $0.1 \%$ contour rather than a local maximum. As we see in Figure
\ref{fig:flame_speed}, the flame settles into a state of steady propagation after an initial
transient period spanning $\sim 3$ ms. The position data here are well fitted by a linear
function of time, and the resulting slope gives the velocity of the flame front.

\begin{deluxetable}{lc}
	\tablecaption{\label{table:flame_speeds_multid} Flame speeds measured in 2D calculations.}
	\tablehead{\colhead{run} & \colhead{$s_\mathrm{front}$ (km s$^{-1}$)}} 
	\startdata
	10/10 & $9.18 \pm 0.03$ \\
	5/5 & $4.00 \pm 0.01$ \\
	10/10-iso7 & $7.56 \pm 0.02$ \\
	10/10-lowres & $9.33 \pm 0.04$ \\
	10/10-1000 Hz & $16.5 \pm 0.21$ \\
	\enddata
	%
\end{deluxetable}

@@

\begin{figure}[t]
\centering
\plotone{speedplot_all.pdf}
\caption{\label{fig:flame_speed}The position of the burning front for each simulation run as a function of time. The dashed lines show linear least squares fits for $t \gtrsim 6$ ms.}
\end{figure}

All of the various approximations have the expected effects. The flame in the 5/5
boost run propagates at about half the speed of the 10/10 flame. This is consistent
with the 1D laminar flames, and is the behavior predicted by Eq.~\ref{eq:flame_scaling}. The
1000~Hz run goes nearly 2 times faster than the 10/10 2000~Hz run, as expected from the
inverse relation between rotation rate and the ratio of burning front area to scale height \citep{cavecchi:2013}.
This confirms that the role of the Coriolis
force is to limit the rate of flame spreading by the anticipated geometrical/hydrodynamical
effect ($s_\mathrm{front} = s_L * L_R / H$). Reducing the resolution had minimal impact on
the flame speed -- the low resolution line is right on top of the fit for the standard run
in Figure \ref{fig:flame_speed}, with their slopes differing by only a few percent. Using a
smaller network also produces similar behavior to the standard run, although there is a small
reduction in speed owing to less energetic burning.

\citet{cavecchi:2013} ran simulations at rotation rates of $450$ Hz exploring a range of opacities.
Accounting for the boosting, we would expect to see our 10/10 flame propagate at $\sim 2.25$ times
the speed of these $450$ Hz runs, but instead the speeds range from $4.5$ to $7$ times those reported in
\citet{cavecchi:2013}. The additional boost may come from cyclic fluid motions that would have been
disallowed in the $450$ Hz runs by the enforcement of vertical hydrostatic equilibrium (we already saw
that hydrodynamic effects can accelerate the burning front in our comparison with the 1D flames). The
inclusion of reaction chains beyond may also increase energy generation and consequently the
propagation speed.

>>>>

%\documentclass[12pt, preprint]{aastex}
\documentclass[trackchanges,preprint,times,tighten]{aastex63}

@@

Small-domain studies of convective burning in XRBs preceding flame
development have been done in two-dimensions \citep{lin:2006,xrb,xrb2}
and three-dimensions \citep{xrb3d}.  These calculations used low Mach
number methods, which approximate the hydrodynamics equations to filter
soundwaves, enabling large timesteps and efficient modeling of
subsonic convection.  While these calculations could not support
the lateral differences needed for flame spreading, they can help
\added{us to} understand the role that convection plays in distributing the initial
burning products vertically throughout the neutron star atmosphere as well
as the nature of any turbulence the burning front might encounter as it propagates
through the atmosphere.

@@

\added{The computational expense of the problem necessitated several approximations
to make this set of simulations \citep{cavecchi:2013, art-2015-cavecchi-etal, art-2016-cavecchi-etal, Cavecchi2019}
tractable. The first was the assumption of vertical hydrostatic equilibrium, which
allowed them to filter out vertically propagating sound waves that would otherwise
have limited the timestep. This enabled them to reach the simulation times (on the
order of seconds) needed to capture the rise of the burst, but they could not model
vortical fluid motions driven by effects such as turbulence and convection. These can
facilitate mixing and help to accelerate the propagation of the flame front. They also
included only the triple-$\alpha$ process in their implementation of nuclear burning,
and use a fixed, constant opacity across the entire domain for each simulation run (although
they explore the dependence of the flame propagation speed on this opacity value).
Our simulations relax the assumption of vertical hydrostatic equilibrium, and implement
a more complete set of nuclear reactions and opacities. However, as discussed later, this
comes at the expense of a different set of approximations.}

@@

The temperature profile in the star and ramp region is given as:
\begin{equation}
T(y) = T_\star + \frac{1}{2} (T_\mathrm{hi} - T_\star) \left [ 1 + \tanh\left( \frac{\tilde{y}}{2 \delta_\mathrm{atm}} \right ) \right ]
\end{equation}
with
\begin{equation}
\tilde{y} = y - H_\star - \frac{3}{2} \delta_\mathrm{atm}
\end{equation}
Here, $\delta_\mathrm{atm}$ is a characteristic width of the
transition ramp, $T_\star$ is the temperature of the neutron star, and
$T_\mathrm{hi}$ is the highest temperature in the HSE model---it will
represent the base of the fuel layer. \added{We also define a floor temperature,
$T_\mathrm{lo}$, which the atmosphere reaches and is held constant at in
the outer, isothermal layer.}

@@

Creating the model involves specifying $T_\star$, $T_\mathrm{hi}$,
$T_\mathrm{lo}$, $\rho_\mathrm{int}$, $H_\star$,
$\delta_\mathrm{atm}$, ${X_k}_\star$, ${X_k}_\mathrm{atm}$, and $g$.  We
then integrate outwards from the base of the ramp region ($y =
H_\star$), enforcing the discrete form of hydrostatic equilibrium,
Eq.~\ref{eq:hse}.  Integrating upwards, we find $p_i$ and
$\rho_i$ using a Newton-Raphson solver together with the equation of state, with either the temperature
specified, $T_i = T(p_i, \rho_i, \{X_k\}_i)$ (in the isothermal, ramp,
and buffer layers) or constant entropy, $s_i = s(p_i, \rho_i,
\{X_k\}_i)$, in the fuel layer.  This follows the procedures described in \citet{ppm-hse}.  We
use a constant temperature for all $y < H_\star +
3\delta_\mathrm{atm}$.  Above this, we switch to an isentropic atmosphere until the
temperature drops to \deleted{a floor value, }$T_\mathrm{lo}$, at which point we
again keep the temperature constant.  The integration of the
atmosphere continues until the density falls to a low density cutoff,
$\rho_\mathrm{cutoff}$.  The material above this height is taken to
have constant density and temperature.
The choice of factors in front of $\delta_\mathrm{atm}$ were designed
to make sure the peak $T$ is attained at the desired density
of the burning layer.
The parameters we use for
the model generation are listed in Table~\ref{table:params} and the 
initial model profiles are showing in Figure~\ref{fig:initial_models}.

@@

To measure the propagation rate of the burning front, we first collapse our nuclear
energy generation rate ($\dot{e}_\mathrm{nuc}$) data at each time into a 1D radial
profile by averaging over the vertical coordinate. We then take the peak
$\dot{e}_\mathrm{nuc}$ value across all profiles to provide a fixed reference point.
We define the position of the flame front to be the location \replaced{ahead}{to the right} of hottest part of
the flame where $\dot{e}_\mathrm{nuc}$ first drops to $< 0.1 \%$ of the global maximum. 
This corresponds roughly to the leading edge of the burning region. Averaging
over the vertical coordinate helps to reduce sensitivity to localized fluid motions, as does
tracking the $0.1 \%$ contour rather than a local maximum. As we see in Figure
\ref{fig:flame_speed}, the flame settles into a state of steady propagation after an initial
transient period spanning $\sim 3$ ms. The position data here are well fitted by a linear
function of time, and the resulting slope gives the velocity of the flame front.

\begin{deluxetable}{lc}
	\tablecaption{\label{table:flame_speeds_multid} Flame speeds measured in 2D calculations.\explain{Updated last entry with some additional data.}}
	\tablehead{\colhead{run} & \colhead{$s_\mathrm{front}$ (km s$^{-1}$)}} 
	\startdata
	10/10 & $9.18 \pm 0.03$ \\
	5/5 & $4.00 \pm 0.01$ \\
	10/10-iso7 & $7.56 \pm 0.02$ \\
	10/10-lowres & $9.33 \pm 0.04$ \\
	10/10-1000 Hz & $16.5 \pm 0.21$ \\
	\enddata
	%
\end{deluxetable}

@@

\begin{figure}[t]
\centering
\plotone{speedplot_all.pdf}
\caption{\label{fig:flame_speed}The position of the burning front for each simulation run as a function of time. The dashed lines show linear least squares fits for $t \gtrsim 6$ ms. \explain{We've updated this with an additional ms of data from the 1000 Hz run.}}
\end{figure}

All of the various approximations have the expected effects. The flame in the 5/5
boost run propagates at about half the speed of the 10/10 flame. This is consistent
with the 1D laminar flames, and is the behavior predicted by Eq.~\ref{eq:flame_scaling}. The
1000~Hz run goes \added{nearly} 2 times faster than the 10/10 2000~Hz run, as expected from the
inverse relation between rotation rate and the ratio of burning front area to scale height
\replaced{\citep{cavecchi:2013}}{(and as observed in \citealt{cavecchi:2013})}.
This confirms that the role of the Coriolis
force is to limit the rate of flame spreading by the anticipated geometrical/hydrodynamical
effect ($s_\mathrm{front} = s_L * L_R / H$). Reducing the resolution
had minimal impact on the flame speed -- the low resolution line is right on top of the fit for the
standard run in Figure \ref{fig:flame_speed}, with their slopes differing by only a few percent. Using a
smaller network also produces similar behavior to the standard run, although there is a small
reduction in speed owing to less energetic burning.

\added{
\citet{cavecchi:2013} ran simulations at rotation rates of $450$ Hz exploring a range of opacities.
Accounting for the boosting, we would expect to see our 10/10 flame propagate at $\sim 2.25$ times
the speed of these $450$ Hz runs, but instead the speeds range from $4.5$ to $7$ times those reported in
\citet{cavecchi:2013}. The additional boost may come from cyclic fluid motions that would have been
disallowed in the $450$ Hz runs by the enforcement of vertical hydrostatic equilibrium (we already saw
that hydrodynamic effects can accelerate the burning front in our comparison with the 1D flames). The
inclusion of reaction chains beyond may also increase energy generation and consequently the
propagation speed.}
